Guilt, shame, and anti-war action in an authoritarian country at war

### 2023-04-27 manuscript.Rmd
Sorted supplementary tables except S1

---
title: "Guilt, shame, and anti-war action in an authoritarian country at war"
output:
  bookdown::pdf_document2:
    latex_engine: lualatex
    keep_tex: yes
    toc: false
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
  - \usepackage{indentfirst}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \usepackage{fontspec}
  - \usepackage{polyglossia}
  - \setdefaultlanguage{english}
  - \setotherlanguages{russian}
  - \setmainfont{Times New Roman}
  - \newfontfamily{\cyrillicfonttt}{Times New Roman} 
  - \usepackage{float} 
  - \usepackage{float} \floatplacement{figure}{H} 
  - \newcommand{\beginsupplement}{\setcounter{table}{0}    
  - \renewcommand{\thetable}{S\arabic{table}} 
  - \setcounter{figure}{0} 
  - \renewcommand{\thefigure}{S\arabic{figure}}}  
indent: true
bibliography: references.bib
csl: elife.csl
---
\vspace{10mm}
\begin{center}
Lusine Grigoryan$^{1}$, Vladimir Ponizovskiy$^{2}$, Marie Isabelle Wießflog$^{2}$, Evgeny Osin$^{3}$,$^{4}$, and Brian Lickel$^{5}$
\vspace{30mm}

@@

$^{4}$ School of Psychology, HSE University
\vspace{5mm}

@@

\noindent All study materials, data, and code can be found on the OSF platform: https://osf.io/4pd2v/?view_only=23597fdc19424706818a72fc0df43009

@@

#national identity
d3 <- d3 %>%
  mutate_at(vars(19:22),~dplyr::recode(., "1\n\nАбсолютно не согласен"=1, 
                                       "2"=2, "3"=3, "4"=4,
                                         "5\nАбсолютно согласен"=5))
describe(d3[,c(19:22)])
psych::alpha(d3[,c(19:22)]) #alpha=.92
d3$NatId <- rowMeans(d3[,c(19:22)], na.rm=TRUE)

@@

describe(d3[,c(28:39)])
psych::alpha(d3[,c(28:33)]) #.95
psych::alpha(d3[,c(34:39)]) #.81

@@

d3$consp5 <- dplyr::recode(d3$consp5, '1'=7,'2'=6,'3'=5,'4'=4,'5'=3,'6'=2,'7'=1)
d3$consp8 <- dplyr::recode(d3$consp8, '1'=7,'2'=6,'3'=5,'4'=4,'5'=3,'6'=2,'7'=1)
cor(d3[,c(56:67)],use="complete.obs")
psych::alpha(d3[,c(56:67)]) #alpha=.89

@@

d3 <- d3 %>%
  mutate_at(vars(68:77),~dplyr::recode(., "абсолютно неверно"=1, 
                                       "едва ли это верно"=2,
                                       "скорее всего - верно"=3,
                                       "совершенно верно"=4))
describe(d3[,c(68:77)])
cor(d3[,c(68:77)],use="complete.obs")
psych::alpha(d3[,c(68:77)]) #alpha=.91

@@

d3 <- d3 %>%
  mutate_at(vars(78:86),~dplyr::recode(., "Совсем не согласен (1)"=1, 
                                       "Не согласен (2)"=2,
                                       "Частично согласен (3)"=3,
                                       "В основном согласен (4)"=4,
                                       "Полностью согласен (5)"=5))
describe(d3[,c(78:86)])
cor(d3[,c(78:86)],use="complete.obs")
psych::alpha(d3[,c(78:86)]) #alpha=.90

@@

cor(d3[,c("sdo1","sdo2","sdo4","sdo6","sdo8")],use="complete.obs")
psych::alpha(d3[,c("sdo1","sdo2","sdo4","sdo6","sdo8")]) #dominance alpha=.75
d3$SDO_d <- rowMeans(d3[,c("sdo1","sdo2","sdo4","sdo6","sdo8")], na.rm=TRUE)

cor(d3[,c("sdo3","sdo5","sdo7","sdo9","sdo10")],use="complete.obs")
psych::alpha(d3[,c("sdo3","sdo5","sdo7","sdo9","sdo10")]) #egalitar. alpha=.77
d3$SDO_e <- rowMeans(d3[,c("sdo3","sdo5","sdo7","sdo9","sdo10")], na.rm=TRUE)

@@

d3 <- d3 %>%
  mutate_at(vars(c("sdo3","sdo5","sdo7","sdo9","sdo10")),~dplyr::recode(., "1"=7, 
                                       "2"=6,"3"=5,"4"=4,"5"=3,"6"=2,
                                       "7"=1))
cor(d3[,c(87:96)],use="complete.obs")
psych::alpha(d3[,c(87:96)]) ##alpha=.80

@@

d3 <- d3 %>%
  mutate_at(vars(97:102),~dplyr::recode(., "1Абсолютно не согласен"=1, 
                                       "2"=2,"3"=3,"4"=4,"5"=5,"6"=6,
                                       "7Абсолютно согласен"=7))
describe(d3[,c(97:102)])
d3$dem3 <- dplyr::recode(d3$dem3, "1"=7,"2"=6,"3"=5,"4"=4,"5"=3,"6"=2,"7"=1)
d3$dem6 <- dplyr::recode(d3$dem6, "1"=7,"2"=6,"3"=5,"4"=4,"5"=3,"6"=2,"7"=1)
cor(d3[,c(97:102)],use="complete.obs")
psych::alpha(d3[,c(97:102)]) #alpha=.69

@@

psych::alpha(d3[,c(103:110)]) #cynicism alpha=.91
psych::alpha(d3[,c(111:114)]) #alienation alpha=.73
psych::alpha(d3[,c(115:119)]) #responsibility alpha=.71

@@

##Attitudes towards the SMO
table(d3$smop_support)
d3$smop_support <- dplyr::recode(d3$smop_support, "Полностью поддерживаю"=5,
                          "Скорее, поддерживаю"=4,
                          "В чём-то поддерживаю, в чём-то нет"=3,
                          "Скорее, не поддерживаю"=2,"Совсем не поддерживаю"=1)
d3$smop_oppos <- dplyr::recode(d3$smop_support,"1"=5,"2"=4,"3"=3,"4"=2,"5"=1)
100*table(d3$smop_support) %>% prop.table() #27.9% oppose, 25.4% not sure, 46.7% support

@@

psych::alpha(d3[,c(121:123)]) #alpha=.90

@@

psych::alpha(d3[,c("cguilt1","cguilt2","cguilt3")]) #alpha=.89
psych::alpha(d3[,c("mshame1","mshame2","mshame3","mshame4")]) #alpha=.93
psych::alpha(d3[,c("ishame1","ishame2","ishame3","ishame4","ishame5")]) #alpha=.94

@@

cor(d3[,c(148:168)],use="complete.obs")
psych::alpha(d3[,c("vib_un1","vib_un2","vib_un3")])
#correlations
#SE:.49, CO:.62, TR:.88, BE:.49, UN: alpha=.75, SD:.63, ST:.66, HE:.57, AC:.67, PO:.52

@@

psych::alpha(d3[,c("motiv1_iR","motiv2_iR","motiv3_i","motiv4_i")], max=10) #alpha = .56

@@

psych::alpha(d3[,c(175:177)]) #alpha for bi_support=.97
psych::alpha(d3[,c(179:181)]) #alpha for bi_oppose=.95

@@

psych::alpha(d3[,c(175:177,253)]) #alpha for future behavior (4 items)=.97
psych::alpha(d3[,c(179:181,252)]) #alpha for future behavior (4 items)=.95

@@

This prediction that guilt would motivate reconciliatory behavior, whereas shame would motivate avoidance is not well supported by the empirical evidence available so far. A recent meta-analysis of group-based emotions ([@Hakim2021], 101 effect sizes, 58 samples, N = 10,305) found that guilt, shame, and anger are equally predictive of support for reparations, with a strong average effect size of r = .44 and no significant differences between emotions. Furthermore, the distinction between moral shame and image shame [@Allpress2010; @Rees2013] helps explain which elements of shame are predictive of reconciliatory behaviors and which are predictive of avoidant behaviors. This distinction is made based on different appraisals that the feeling of shame can be accompanied with: we feel moral shame when the group members’ behavior violates important moral standards or values of the group, and we feel image shame when the group members’ behavior tarnishes the ingroup’s reputation [@Rees2013]. Studies that make the distinction between moral shame and image shame find that moral shame is predictive of reconciliatory behaviors on par or more strongly than guilt, whereas image shame is predictive of distancing and avoidance  [@Allpress2010; @Allpress2014; @Grigoryan2017; @Nooitgedagt2021; @Rees2013]. We therefore expect that moral shame and guilt, but not image shame, will predict a stronger intention to act against the war in Russia (H1).

@@

We collected a quota sample, representative of the Russian population by gender and age. Participants were recruited via the online crowdsourcing platform Yandex Toloka in August 2022.  This was the only feasible option to collect fully anonymized data in Russia, which was crucial, given that the repercussions for expressing opposition to the war can be severe. Participants received \$ 0.5 (30 RUB) for their participation. A total of N=`r length(d2$start)` participants gave full informed consent to participate in the study. Thirty participants were excluded since they dropped out of the study before completion, and another eight were excluded for failing more than one out of three attention checks. The effective sample size is N=`r length(d3$start)`.

@@

<!-- check dollar/percentages/mean ect. -->

@@

#mean responses on behavioral intention questionnaires for each participant stored in means_bi 
means_bi <- NULL
for (i in 1:length(d3$bi_smop_opp1)) {
  means_bi[i] <- round((d3$bi_smop_opp1[i] + d3$bi_smop_opp2[i] + d3$bi_smop_opp3[i])/3, digits = 0)
}
```

***Anti-war actions***. We measured past behavior by asking how many of the eight political actions listed did the participant do to oppose the war in Ukraine (e.g., “signing a petition”, “participating in a demonstration”, etc.). The responses ranged from 0 to 8, with `r round(table(d3$pb_smop_opp)[1]/length(d3$pb_smop_opp)*100, digits = 0)`% of participants reporting zero actions. Behavioral intention to participate in any political action against the war was measured with 3 items (“I want to/intend to/plan to take part in political action against the war in Ukraine”), on a 5-point scale from 1 – ‘ Completely disagree’ to 5 – ‘Completely agree’ ($\alpha$ = .95; `r round(table(means_bi)[1]/length(d3$bi_smop_opp1)*100, digits = 0)`% of participants had a mean score of 1). Finally, we asked about the probability that the person will take any political action, from 1 – highly improbable, 10 – (almost) certainly (`r round(table(d3$bp_smop_opp)[1]/length(d3$bp_smop_opp)*100, digits = 0)`% answered ‘1’).

***Attitude towards the war*** was measured with 3 items, on a bipolar 5-pt Osgood-style scale: “The special military operation in Ukraine is…”: bad-good, harmful-beneficial, useless-useful ($\alpha$ = .90, M = `r round(mean(d3$smop_att, na.rm = TRUE), digits = 2)`, SD = `r round(sd(d3$smop_att, na.rm = TRUE), digits = 2)`).

@@

*Support for democratic values* was measured with a 6-item scale from @CanettiNisim2007. Example item: "Every citizen has the right to take his convictions to the street if necessary" ($\alpha$ = .69).

 *Political alienation* was measured with four items from @Olsen1969. Example item: "There is not much that people like me can do to influence actions of the government" ($\alpha$ = .73).

*Political cynicism* was measured with eight items from @Pattyn2012. Example item: "Politicians pretend to care more about people than they really do" ($\alpha$ = .91).

*Political responsibility.* We formulated five items aimed at capturing participants’ beliefs about political responsibility: “The government will face difficulties if it tries to do something that most people in the country do not agree with”, “Participation in political life is a responsibility of every person before themselves and their fellow countrymen”, “If I see that the country is going in the wrong direction, it is my duty to get my voice heard by the government”, “The citizens of Russia are accountable for the actions of their government”, “If the majority of the citizens are against some decision made by the government, they can influence that decision” ($\alpha$ = .71).

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = `r round(summary5$fit[9], digits = 3)`, RMSEA = `r round(summary5$fit[17], digits = 3)`, SRMR = `r round(summary5$fit[25], digits = 3)`). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = `r round(summary6$fit[9], digits = 3)`, RMSEA = `r round(summary6$fit[17], digits = 3)`, SRMR = `r round(summary6$fit[25], digits = 3)`). This modified model includes the political responsibility factor with 5 items, political alienation factor with 3 items (the item “I believe public officials don't care much what people like me think” was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: “politicians are evil” (4 items, e.g., “Politicians are only interested in getting and maintaining power”) and “politics is evil” (4 items, e.g., “No man can hope to stay honest once he enters politics”).

***Positive and negative emotions.*** We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale [@Watson1988]. Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-pt scale (1 – ‘do not experience at all’ to 5 – ‘extremely’). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt and/or shame, respectively (we took this precaution because of negative prior experiences when using these measures in Russia; see @Grigoryan2018; @Grigoryan2017, for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame) using CFA and found the fit to be unsatisfactory (CFI = `r round(summary7$fit[9], digits = 3)`, RMSEA = `r round(summary7$fit[17], digits = 3)`, SRMR = `r round(summary7$fit[25], digits = 3)`). A 3-factor structure with positive emotions (joy, hope, enthusiasm), negative dominant emotions (anger, disgust, contempt) and negative submissive emotions (fear, sadness, depression) showed a satisfactory fit to the data (CFI = `r round(summary8$fit[9], digits = 3)`, RMSEA = `r round(summary8$fit[17], digits = 3)`, SRMR = `r round(summary8$fit[25], digits = 3)`) [see @Mehrabian1980], for the distinction between dominant and submissive emotions).

***Group-based guilt and shame.*** We used the measures of group-based guilt, image shame, and moral shame developed by @Rees2013, Russian adaptation by @Grigoryan2017. All items are answered on a 5-pt scale from 1 – ‘Completely disagree’ to 5 – ‘Completely agree’. Guilt was measured with 3 items (e.g., “I feel guilty for the manner in which Ukrainian people have been treated by Russians”, $\alpha$ = .89), image shame with 5 items (“I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine”, $\alpha$ = .94), and moral shame with 4 items (“I feel ashamed because Russia’s actions with regard to Ukraine have been immoral”, $\alpha$ = .93). All missing values were replaced with zeros, since participants who didn’t see the scale were the ones who reported no feelings of guilt or shame in the PANAS.

@@

##past behavior: linear (Table 2)
table(d3$pb_smop_opp)
h3a <- lm(pb_smop_opp ~ cguiltf+ishamef+mshamef,data = d3)
summary(h3a)
h3a2 <- lm(scale(pb_smop_opp) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef),data = d3)
summary(h3a)

##past behavior: z-i negbin (Table S3)
h3a3 <- zeroinfl(pb_smop_opp ~ cguiltf+ishamef+mshamef,data = d3,dist="negbin")
summary(h3a)

##behavioral intention (3 items): linear (Table 2)
table(d3$bi_oppos)
h3b <- lm(bi_oppos ~ cguiltf+ishamef+mshamef,data = d3)
summary(h3b)
h3b2 <- lm(scale(bi_oppos) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef),data = d3)
summary(h3b)

##behavioral intention: z-i negbin (Table S3)
d3$bi_opp0 <- d3$bi_oppos-1
d3$bi_opp0 <- round(d3$bi_opp0)
table(d3$bi_opp0)
h3b3 <- zeroinfl(bi_opp0 ~ cguiltf+ishamef+mshamef, data = d3,dist="negbin")
summary(h3b)

##probability of doing something in the future: linear (Table 2)
table(d3$bp_smop_opp)
h3c <- lm(bp_smop_opp ~ cguiltf+ishamef+mshamef, data = d3)
h3coutput <- summary(h3c)
h3c2 <- lm(scale(bp_smop_opp) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef), data = d3)
summary(h3c)

##probability of doing something in the future: z-i negbin (Table S3)
d3$bp_smop_opp0 <- d3$bp_smop_opp-1
table(d3$bp_smop_opp0)
h3c3 <- zeroinfl(bp_smop_opp0 ~ cguiltf+ishamef+mshamef, data = d3,dist="negbin")
summary(h3c)

@@

d3$fb_oppos_count <- rowSums(d3[,c("bi_smop_opp1R","bi_smop_opp2R","bi_smop_opp3R",
                                   "bp_smop_oppR")],na.rm=T)
table(d3$fb_oppos_count)
100*table(d3$fb_oppos_count)%>% prop.table()

@@

>>>>

---
title: "Guilt, shame, and anti-war action in an authoritarian country at war"
output:
  bookdown::pdf_document2:
    latex_engine: lualatex
    keep_tex: yes
    toc: false
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
  - \usepackage{indentfirst}
  - \usepackage{sectsty}
  - \allsectionsfont{\centering}
  - \usepackage{fontspec}
  - \usepackage{polyglossia}
  - \setdefaultlanguage{english}
  - \setotherlanguages{russian}
  - \setmainfont{Times New Roman}
  - \newfontfamily{\cyrillicfonttt}{Times New Roman} 
  - \usepackage{float} 
  - \usepackage{float} \floatplacement{figure}{H} 
  - \newcommand{\beginsupplement}{\setcounter{table}{0}    
  - \renewcommand{\thetable}{S\arabic{table}} 
  - \setcounter{figure}{0} 
  - \renewcommand{\thefigure}{S\arabic{figure}}}  
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
indent: true
bibliography: references.bib
csl: elife.csl
---

\fancyhead[LH]{GUILT, SHAME, AND ANTI-WAR ACTION}

\vspace{10mm}
\begin{center}
Lusine Grigoryan$^{1}$, Vladimir Ponizovskiy$^{2}$, Marie Isabelle Wießflog$^{2}$, Evgeny Osin$^{3}$,$^{4}$, and Brian Lickel$^{5}$
\vspace{30mm}

@@

$^{4}$ Department of Psychology, HSE University
\vspace{5mm}

@@

\noindent All study materials, data, and code can be found on the OSF platform:

\noindent https://osf.io/4pd2v/?view_only=23597fdc19424706818a72fc0df43009

@@

#national identity
d3 <- d3 %>%
  mutate_at(vars(19:22),~dplyr::recode(., "1\n\nАбсолютно не согласен"=1, 
                                       "2"=2, "3"=3, "4"=4,
                                         "5\nАбсолютно согласен"=5))
describe(d3[,c(19:22)])
alpha_NatID <- psych::alpha(d3[,c(19:22)]) #alpha=.92
d3$NatId <- rowMeans(d3[,c(19:22)], na.rm=TRUE)

@@

describe(d3[,c(28:39)])
alpha_bcp1 <- psych::alpha(d3[,c(28:33)]) #.95
alpha_bcp2 <- psych::alpha(d3[,c(34:39)]) #.81

@@

d3$consp5 <- dplyr::recode(d3$consp5, '1'=7,'2'=6,'3'=5,'4'=4,'5'=3,'6'=2,'7'=1)
d3$consp8 <- dplyr::recode(d3$consp8, '1'=7,'2'=6,'3'=5,'4'=4,'5'=3,'6'=2,'7'=1)
cor(d3[,c(56:67)],use="complete.obs")
alpha_consp <- psych::alpha(d3[,c(56:67)]) #alpha=.89

@@

d3 <- d3 %>%
  mutate_at(vars(68:77),~dplyr::recode(., "абсолютно неверно"=1, 
                                       "едва ли это верно"=2,
                                       "скорее всего - верно"=3,
                                       "совершенно верно"=4))
describe(d3[,c(68:77)])
cor(d3[,c(68:77)],use="complete.obs")
alpha_seff <- psych::alpha(d3[,c(68:77)]) #alpha=.91

@@

d3 <- d3 %>%
  mutate_at(vars(78:86),~dplyr::recode(., "Совсем не согласен (1)"=1, 
                                       "Не согласен (2)"=2,
                                       "Частично согласен (3)"=3,
                                       "В основном согласен (4)"=4,
                                       "Полностью согласен (5)"=5))
describe(d3[,c(78:86)])
cor(d3[,c(78:86)],use="complete.obs")
alpha_RWA <- psych::alpha(d3[,c(78:86)]) #alpha=.90

@@

cor(d3[,c("sdo1","sdo2","sdo4","sdo6","sdo8")],use="complete.obs")
alpha_SDO_d <- psych::alpha(d3[,c("sdo1","sdo2","sdo4","sdo6","sdo8")]) #dominance alpha=.75
d3$SDO_d <- rowMeans(d3[,c("sdo1","sdo2","sdo4","sdo6","sdo8")], na.rm=TRUE)

cor(d3[,c("sdo3","sdo5","sdo7","sdo9","sdo10")],use="complete.obs")
alpha_SDO_e <- psych::alpha(d3[,c("sdo3","sdo5","sdo7","sdo9","sdo10")]) #egalitar. alpha=.77
d3$SDO_e <- rowMeans(d3[,c("sdo3","sdo5","sdo7","sdo9","sdo10")], na.rm=TRUE)

@@

d3 <- d3 %>%
  mutate_at(vars(c("sdo3","sdo5","sdo7","sdo9","sdo10")),~dplyr::recode(., "1"=7, 
                                       "2"=6,"3"=5,"4"=4,"5"=3,"6"=2,
                                       "7"=1))
cor(d3[,c(87:96)],use="complete.obs")
alpha_SDO <- psych::alpha(d3[,c(87:96)]) ##alpha=.80

@@

d3 <- d3 %>%
  mutate_at(vars(97:102),~dplyr::recode(., "1Абсолютно не согласен"=1, 
                                       "2"=2,"3"=3,"4"=4,"5"=5,"6"=6,
                                       "7Абсолютно согласен"=7))
describe(d3[,c(97:102)])
d3$dem3 <- dplyr::recode(d3$dem3, "1"=7,"2"=6,"3"=5,"4"=4,"5"=3,"6"=2,"7"=1)
d3$dem6 <- dplyr::recode(d3$dem6, "1"=7,"2"=6,"3"=5,"4"=4,"5"=3,"6"=2,"7"=1)
cor(d3[,c(97:102)],use="complete.obs")
alpha_demv <- psych::alpha(d3[,c(97:102)]) #alpha=.69

@@

alpha_cyn <- psych::alpha(d3[,c(103:110)]) #cynicism alpha=.91
alpha_alien <- psych::alpha(d3[,c(111:114)]) #alienation alpha=.73
alpha_resp <- psych::alpha(d3[,c(115:119)]) #responsibility alpha=.71

@@

##Attitudes towards the SMO
table(d3$smop_support)
d3$smop_support <- dplyr::recode(d3$smop_support, "Полностью поддерживаю"=5,
                          "Скорее, поддерживаю"=4,
                          "В чём-то поддерживаю, в чём-то нет"=3,
                          "Скорее, не поддерживаю"=2,"Совсем не поддерживаю"=1)
d3$smop_oppos <- dplyr::recode(d3$smop_support,"1"=5,"2"=4,"3"=3,"4"=2,"5"=1)
smotable <- 100*table(d3$smop_support) %>% prop.table() #27.9% oppose, 25.4% not sure, 46.7% support

@@

alpha_SMO <- psych::alpha(d3[,c(121:123)]) #alpha=.90

@@

alpha_guilt <- psych::alpha(d3[,c("cguilt1","cguilt2","cguilt3")]) #alpha=.89
alpha_mshame <- psych::alpha(d3[,c("mshame1","mshame2","mshame3","mshame4")]) #alpha=.93
alpha_ishame <- psych::alpha(d3[,c("ishame1","ishame2","ishame3","ishame4","ishame5")]) #alpha=.94

@@

cor(d3[,c(148:168)],use="complete.obs")
alpha_vib <- psych::alpha(d3[,c("vib_un1","vib_un2","vib_un3")])
#correlations
#SE:.49, CO:.62, TR:.88, BE:.49, UN: alpha=.75, SD:.63, ST:.66, HE:.57, AC:.67, PO:.52

@@

alpha_motiv <- psych::alpha(d3[,c("motiv1_iR","motiv2_iR","motiv3_i","motiv4_i")], max=10) #alpha = .56

@@

alpha_bi_supp <- psych::alpha(d3[,c(175:177)]) #alpha for bi_support=.97
alpha_bi_oppos <- psych::alpha(d3[,c(179:181)]) #alpha for bi_oppose=.95

@@

alpha_fb_support <- psych::alpha(d3[,c(175:177,253)]) #alpha for future behavior (4 items)=.97
alpha_fb_oppose <- psych::alpha(d3[,c(179:181,252)]) #alpha for future behavior (4 items)=.95

@@

This prediction that guilt would motivate reconciliatory behavior, whereas shame would motivate avoidance is not well supported by the empirical evidence available so far. A recent meta-analysis of group-based emotions ([@Hakim2021], 101 effect sizes, 58 samples, N = 10,305) found that guilt, shame, and anger are equally predictive of support for reparations, with a strong average effect size of *r* = .44 and no significant differences between emotions. Furthermore, the distinction between moral shame and image shame [@Allpress2010; @Rees2013] helps explain which elements of shame are predictive of reconciliatory behaviors and which are predictive of avoidant behaviors. This distinction is made based on different appraisals that the feeling of shame can be accompanied with: we feel moral shame when the group members’ behavior violates important moral standards or values of the group, and we feel image shame when the group members’ behavior tarnishes the ingroup’s reputation [@Rees2013]. Studies that make the distinction between moral shame and image shame find that moral shame is predictive of reconciliatory behaviors on par or more strongly than guilt, whereas image shame is predictive of distancing and avoidance  [@Allpress2010; @Allpress2014; @Grigoryan2017; @Nooitgedagt2021; @Rees2013]. We therefore expect that moral shame and guilt, but not image shame, will predict a stronger intention to act against the war in Russia (H1).

@@

We collected a quota sample, representative of the Russian population by gender and age. Participants were recruited via the online crowdsourcing platform Yandex Toloka in August 2022.  This was the only feasible option to collect fully anonymized data in Russia, which was crucial, given that the repercussions for expressing opposition to the war can be severe. Participants received \$0.5 (30 RUB) for their participation. A total of N=`r length(d2$start)` participants gave full informed consent to participate in the study. Thirty participants were excluded since they dropped out of the study before completion, and another eight were excluded for failing more than one out of three attention checks. The effective sample size is N=`r length(d3$start)`.

@@

@@

#mean responses on behavioral intention questionnaires for each participant stored in means_bi 
means_bi <- NULL
for (i in 1:length(d3$bi_smop_opp1)) {
  means_bi[i] <- round((d3$bi_smop_opp1[i] + d3$bi_smop_opp2[i] + d3$bi_smop_opp3[i])/3, digits = 0)
}

# some slight mismatches with figures in the paragraph below, 83 vs 83% asnd 73 vs 71%

***Anti-war actions***. We measured *past behavior* by asking how many of the eight political actions listed did the participant do to oppose the war in Ukraine (e.g., “signing a petition”, “participating in a demonstration”, etc.). The responses ranged from 0 to 8, with `r round(table(d3$pb_smop_opp)[1]/length(d3$pb_smop_opp)*100, digits = 0)`% of participants reporting zero actions. *Behavioral intention* to participate in any political action against the war was measured with 3 items (“I want to/intend to/plan to take part in political action against the war in Ukraine”), on a 5-point scale from 1 – ‘ Completely disagree’ to 5 – ‘Completely agree’ ($\alpha$ = `r round(alpha_bi_oppos$total[1],digits=2)`; `r round(table(means_bi)[1]/length(d3$bi_smop_opp1)*100, digits = 0)`% of participants had a mean score of 1). Finally, we asked about *the probability that the person will take any political action*, from 1 – highly improbable, 10 – (almost) certainly (`r round(table(d3$bp_smop_opp)[1]/length(d3$bp_smop_opp)*100, digits = 0)`% answered ‘1’).

***Attitude towards the war*** was measured with 3 items, on a bipolar 5-pt Osgood-style scale: “The special military operation in Ukraine is…”: bad-good, harmful-beneficial, useless-useful ($\alpha$ = `r round(alpha_SMO$total[1],digits=2)`, M = `r round(mean(d3$smop_att, na.rm = TRUE), digits = 2)`, SD = `r round(sd(d3$smop_att, na.rm = TRUE), digits = 2)`).

@@

*Support for democratic values* was measured with a 6-item scale from @CanettiNisim2007. Example item: "Every citizen has the right to take his convictions to the street if necessary" ($\alpha$ = `r round(alpha_demv$total[1],digits=2)`).

*Political alienation* was measured with four items from @Olsen1969. Example item: "There is not much that people like me can do to influence actions of the government" ($\alpha$ = `r round(alpha_alien$total[1],digits=2)`).

*Political cynicism* was measured with eight items from @Pattyn2012. Example item: "Politicians pretend to care more about people than they really do" ($\alpha$ = `r round(alpha_cyn$total[1],digits=2)`).

*Political responsibility.* We formulated five items aimed at capturing participants’ beliefs about political responsibility: “The government will face difficulties if it tries to do something that most people in the country do not agree with”, “Participation in political life is a responsibility of every person before themselves and their fellow countrymen”, “If I see that the country is going in the wrong direction, it is my duty to get my voice heard by the government”, “The citizens of Russia are accountable for the actions of their government”, “If the majority of the citizens are against some decision made by the government, they can influence that decision” ($\alpha$ = `r round(alpha_resp$total[1],digits=2)`).

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = `r round(summary5$fit[9], digits = 3)`, RMSEA = `r round(summary5$fit[17], digits = 3)`, SRMR = `r round(summary5$fit[25], digits = 3)`). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = `r round(summary6$fit[9], digits = 3)`, RMSEA = `r round(summary6$fit[17], digits = 3)`, SRMR = `r round(summary6$fit[25], digits = 3)`). This modified model includes the *political responsibility* factor with 5 items, *political alienation* factor with 3 items (the item “I believe public officials don't care much what people like me think” was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: “*politicians are evil*” (4 items, e.g., “Politicians are only interested in getting and maintaining power”) and “*politics is evil*” (4 items, e.g., “No man can hope to stay honest once he enters politics”).

***Positive and negative emotions.*** We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale [@Watson1988]. Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-pt scale (1 – ‘do not experience at all’ to 5 – ‘extremely’). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt and/or shame, respectively (we took this precaution because of negative prior experiences when using these measures in Russia; see @Grigoryan2018; @Grigoryan2017, for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame) using CFA and found the fit to be unsatisfactory (CFI = `r round(summary7$fit[9], digits = 3)`, RMSEA = `r round(summary7$fit[17], digits = 3)`, SRMR = `r round(summary7$fit[25], digits = 3)`). A 3-factor structure with positive emotions (joy, hope, enthusiasm), negative dominant emotions (anger, disgust, contempt) and negative submissive emotions (fear, sadness, depression) showed a satisfactory fit to the data (CFI = `r round(summary8$fit[9], digits = 3)`, RMSEA = `r round(summary8$fit[17], digits = 3)`, SRMR = `r round(summary8$fit[25], digits = 3)`) (see @Mehrabian1980, for the distinction between dominant and submissive emotions).

***Group-based guilt and shame.*** We used the measures of group-based guilt, image shame, and moral shame developed by @Rees2013, Russian adaptation by @Grigoryan2017. All items are answered on a 5-pt scale from 1 – ‘Completely disagree’ to 5 – ‘Completely agree’. Guilt was measured with 3 items (e.g., “I feel guilty for the manner in which Ukrainian people have been treated by Russians”, $\alpha$ = `r round(alpha_guilt$total[1],digits=2)`), image shame with 5 items (“I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine”, $\alpha$ = `r round(alpha_ishame$total[1],digits=2)`), and moral shame with 4 items (“I feel ashamed because Russia’s actions with regard to Ukraine have been immoral”, $\alpha$ = `r round(alpha_mshame$total[1],digits=2)`). All missing values were replaced with zeros, since participants who didn’t see the scale were the ones who reported no feelings of guilt or shame in the PANAS.

@@

##past behavior: linear (Table 2)
table(d3$pb_smop_opp)
h3a <- lm(pb_smop_opp ~ cguiltf+ishamef+mshamef,data = d3)
summary(h3a)
h3a2 <- lm(scale(pb_smop_opp) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef),data = d3)
summary(h3a2)

##past behavior: z-i negbin (Table S3)
h3a3 <- zeroinfl(pb_smop_opp ~ cguiltf+ishamef+mshamef,data = d3,dist="negbin")
summary(h3a3)

##behavioral intention (3 items): linear (Table 2)
table(d3$bi_oppos)
h3b <- lm(bi_oppos ~ cguiltf+ishamef+mshamef,data = d3)
summary(h3b)
h3b2 <- lm(scale(bi_oppos) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef),data = d3)
summary(h3b2)

##behavioral intention: z-i negbin (Table S3)
d3$bi_opp0 <- d3$bi_oppos-1
d3$bi_opp0 <- round(d3$bi_opp0)
table(d3$bi_opp0)
h3b3 <- zeroinfl(bi_opp0 ~ cguiltf+ishamef+mshamef, data = d3,dist="negbin")
summary(h3b3)

##probability of doing something in the future: linear (Table 2)
table(d3$bp_smop_opp)
h3c <- lm(bp_smop_opp ~ cguiltf+ishamef+mshamef, data = d3)
h3coutput <- summary(h3c)
h3c2 <- lm(scale(bp_smop_opp) ~ scale(cguiltf)+scale(ishamef)+scale(mshamef), data = d3)
summary(h3c2)

##probability of doing something in the future: z-i negbin (Table S3)
d3$bp_smop_opp0 <- d3$bp_smop_opp-1
table(d3$bp_smop_opp0)
h3c3 <- zeroinfl(bp_smop_opp0 ~ cguiltf+ishamef+mshamef, data = d3,dist="negbin")
summary(h3c3)

@@

d3$fb_oppos_count <- rowSums(d3[,c("bi_smop_opp1R","bi_smop_opp2R","bi_smop_opp3R",
                                   "bp_smop_oppR")],na.rm=T)
table(d3$fb_oppos_count)
h5counts <- 100*table(d3$fb_oppos_count)%>% prop.table()

@@

```{r TableS1}
